## Study Design and Algorithmic Bias

The principle "garbage in, garbage out" remains fundamental in data science. This exercise demonstrates how even well-implemented models cannot overcome the limitations of poor-quality data—a challenge that becomes increasingly significant in our current data landscape.

### Objectives

Our goal is this exercise is to illustrate how flaws in sampling design direct affect model predictions in ways that can perpetuate or amplify existing biases in various algorithmics. 


### Procedure

1. **Data Acquisition**
   
   Download the dataset `Loan_Application_Data_subset.csv`, containing 10,000 randomly selected records from the [2023 FFIEC national loan-level dataset](https://ffiec.cfpb.gov/data-publication/dynamic-national-loan-level-dataset/2023), which encompasses national-level data on all home loans issued in the United States during 2023.

2. **Data Partitioning**
   
   First remove any loans that were not accepted:
   - A loan is denoted as accepted if it has loan_data$action_taken == 1
   
   Divide the dataset into two segments:
   - Training set: First 3,000 loan records
   - Test set: Remaining 1,955 loan records
   
   The training set will be used to develop a deliberately biased model to demonstrate how this bias affects subsequent analysis.
   
```{r}
# Load the dataset
loan_data <- read.csv("Loan_Application_Data_subset.csv")

loan_data = loan_data[loan_data$action_taken == 1,]

str(loan_data)

# View the first few rows
head(loan_data)

# Summary statistics
summary(loan_data)

# Check for missing values
colSums(is.na(loan_data))
```
   

3. **Initial Data Analysis**
   
   Before manipulation, examine the dataset using linear regression and the LASSO model provided in `Lending_LASSO.R`. 
   
   -Make a model that predicts loan amount [loan_amount] based on income [income], interest rate [interest_rate], sex [derived_sex], and race [derived_race] of the applicant.
   - Analyze the relationships between these predictor variables and loan amount
   - Compare LASSO results with standard linear regression outputs
   
```{r data_partition}
# Create training and test sets
train_data <- loan_data[1:3000, ]
test_data <- loan_data[3001:4955, ]
train_data = train_data[,c('action_taken','loan_amount','income','interest_rate','derived_sex','derived_race')]
train_data = train_data[complete.cases(train_data),]
test_data = test_data[,c('action_taken','loan_amount','income','interest_rate','derived_sex','derived_race')]
test_data = test_data[complete.cases(test_data),]

train_data$interest_rate = as.numeric(train_data$interest_rate)
test_data$interest_rate = as.numeric(test_data$interest_rate)


# Verify the dimensions
dim(train_data)
dim(test_data)
```

4. **Creating a Biased Dataset**
   
   To mimic the effect of a biased sample design, create a new dataset from the training data which oversamples low income female. For each datapoints in this biased sample:
   - Include female applicants with less than the median loan amount with 80% probability
   - Include non-female applicants (male, joint, or undisclosed) with 20% probability
   
   
   
```{r create_biased_data}
# Function to create biased dataset with specified sampling probabilities
create_biased_data <- function(data, female_prob = 0.8, non_female_prob = 0.2) {
  set.seed(456)
  
  # Create separate subsets for female and non-female applicants
  female_applicants <- data[((data$derived_sex == "Female")&(data$loan_amount <= 225000)), ]
  non_female_applicants <- data[!(((data$derived_sex == "Female")&(data$loan_amount <= 225000))), ]
  
  # Sample with specified probabilities
  female_sample <- female_applicants[sample(nrow(female_applicants), 
                                           size = round(nrow(female_applicants) * female_prob)), ]
  
  non_female_sample <- non_female_applicants[sample(nrow(non_female_applicants), 
                                                    size = round(nrow(non_female_applicants) * non_female_prob)), ]
  
  # Combine the samples to create the biased dataset
  biased_data <- rbind(female_sample, non_female_sample)
  
  # Shuffle the rows to mix the data
  biased_data <- biased_data[sample(nrow(biased_data)), ]
  
  return(biased_data)
}


biased_train_data = create_biased_data(train_data)
biased_train_data = biased_train_data[complete.cases(biased_train_data),]
```

5. **Model Comparison**
   
   - Fit a linear regression to both the biased and unbiased training dataset.
   - Generate loan amount predictions using the biased and unbiased linear regressions
   - Compare the predictions made using the biased and unbiased linear regressions. Are there any patterns in the errors?
   - Analyze prediction accuracy stratified by applicant demographics and identify whether certain groups receive better or worse estimates
   -If time, repeat the above with LASSO and random Forest
   
```{r model_comparison}
# Create model matrix for LASSO on biased data
x_biased <- model.matrix(~ income + interest_rate + factor(derived_sex) + factor(derived_race) , data = biased_train_data)
y_biased <- biased_train_data$loan_amount

#linear regression
lm_model_biased <- lm(log(loan_amount) ~ income + interest_rate + factor(derived_sex), data = biased_train_data)


#linear regression
lm_model_unbiased <- lm(log(loan_amount) ~ income + interest_rate + factor(derived_sex) , data = train_data)


```

#Random forest 
```{r}
library(randomForest)
rf_biased <- randomForest(log(loan_amount)~., data=biased_train_data, proximity=TRUE)

rf_unbiased <- randomForest(log(loan_amount)~., data=train_data[complete.cases(train_data),], proximity=TRUE)


```


   

```{r}
#x_test <- model.matrix(~ income + interest_rate + factor(derived_sex) , data = test_data)

y_test <- log(test_data$loan_amount)
y_hat = (predict(lm_model_unbiased,newdata = test_data))
y_hat_biased =  (predict(lm_model_biased,newdata = test_data))


plot(y_test,y_hat)
abline(a=0,b=1)

plot(y_test,y_hat_biased)
abline(a=0,b=1)

plot(y_test[test_data$derived_sex == 'Female'],y_hat[test_data$derived_sex == 'Female'], ylim = c(9,14), xlim = c(9,14))
abline(a=0,b=1)

plot(y_test[test_data$derived_sex == 'Female'],y_hat_biased[test_data$derived_sex == 'Female'], ylim = c(9,14), xlim = c(9,14))
abline(a=0,b=1)

plot(y_hat[test_data$derived_sex == 'Female'], y_hat_biased[test_data$derived_sex == 'Female'], ylim = c(11,14), xlim = c(11,14))
abline(a=0,b=1)

plot(y_hat[test_data$derived_sex == 'Male'], y_hat_biased[test_data$derived_sex == 'Male'], ylim = c(11,14), xlim = c(11,14))
abline(a=0,b=1)

plot(y_hat[test_data$derived_sex == 'Joint'], y_hat_biased[test_data$derived_sex == 'Joint'], ylim = c(11,14), xlim = c(11,14))
abline(a=0,b=1)

plot(y_hat[test_data$derived_sex == 'Sex Not Available'], y_hat_biased[test_data$derived_sex == 'Sex Not Available'], ylim = c(11,14), xlim = c(11,14))
abline(a=0,b=1)



```

```{r}
y_test <- log(test_data$loan_amount)
y_hat = (predict(rf_unbiased,newdata = test_data))
y_hat_biased =  (predict(rf_biased,newdata = test_data))


plot(y_test,y_hat)
abline(a=0,b=1)

plot(y_test,y_hat_biased)
abline(a=0,b=1)

plot(y_test[test_data$derived_sex == 'Female'],y_hat[test_data$derived_sex == 'Female'], ylim = c(9,14), xlim = c(9,14))
abline(a=0,b=1)

plot(y_test[test_data$derived_sex == 'Female'],y_hat_biased[test_data$derived_sex == 'Female'], ylim = c(9,14), xlim = c(9,14))
abline(a=0,b=1)

plot(y_hat[test_data$derived_sex == 'Female'], y_hat_biased[test_data$derived_sex == 'Female'], ylim = c(11,14), xlim = c(11,14))
abline(a=0,b=1)

plot(y_hat[test_data$derived_sex == 'Male'], y_hat_biased[test_data$derived_sex == 'Male'], ylim = c(11,14), xlim = c(11,14))
abline(a=0,b=1)

plot(y_hat[test_data$derived_sex == 'Joint'], y_hat_biased[test_data$derived_sex == 'Joint'], ylim = c(11,14), xlim = c(11,14))
abline(a=0,b=1)

plot(y_hat[test_data$derived_sex == 'Sex Not Available'], y_hat_biased[test_data$derived_sex == 'Sex Not Available'], ylim = c(11,14), xlim = c(11,14))
abline(a=0,b=1)



```


```{r}

plot(y_hat[test_data$derived_sex == 'Male'], y_hat_biased[test_data$derived_sex == 'Male'])
abline(a=0,b=1)

plot(y_hat[test_data$derived_sex == 'Joint'], y_hat_biased[test_data$derived_sex == 'Joint'])
abline(a=0,b=1)

plot(y_hat[test_data$derived_sex == 'Sex Not Available'], y_hat_biased[test_data$derived_sex == 'Sex Not Available'])
abline(a=0,b=1)
```

   