## The Impact of Data Bias on Model Performance

The principle "garbage in, garbage out" remains fundamental in data science. This exercise demonstrates how even well-implemented models cannot overcome the limitations of poor-quality data—a challenge that becomes increasingly significant in our current data landscape dominated by synthetically generated information.

### Objectives

This exercise will illustrate how sampling bias affects model predictions and can perpetuate or amplify existing biases in lending data.

### Procedure

1. **Data Acquisition**
   
   Download the dataset `Loan_Application_Data_subset.csv`, containing 10,000 randomly selected records from the [2023 FFIEC national loan-level dataset](https://ffiec.cfpb.gov/data-publication/dynamic-national-loan-level-dataset/2023), which encompasses national-level data on all home loans issued in the United States during 2023.

2. **Data Partitioning**
   
   Divide the dataset into two segments:
   - Training set: First 8,000 loan records
   - Test set: Remaining 2,000 loan records
   
   The training set will be used to develop a deliberately biased model to demonstrate how this bias affects subsequent analysis.
   
```{r load_data}
# Load the dataset
loan_data <- read.csv("Loan_Application_Data_subset.csv")

# Display the structure of the dataset
str(loan_data)

# View the first few rows
head(loan_data)

# Summary statistics
summary(loan_data)

# Check for missing values
colSums(is.na(loan_data))
```
   

3. **Initial Data Analysis**
   
   Before manipulation, examine the dataset using the LASSO model provided in `Lending_LASSO.R`. This model predicts loan amount [loan_amount] based on income [income], interest rate [interest_rate], sex [derived_sex], and race [derived_race] of the applicant.
   
   - Analyze the relationships between these predictor variables and loan amount
   - Compare LASSO results with standard linear regression outputs
   - Note that this simplified model excludes several important factors, so all analyses should be considered preliminary
   
```{r data_partition}
loan_data$approved = loan_data$action_taken == 2
# Create training and test sets
train_data <- loan_data[1:8000, ]
test_data <- loan_data[8001:10000, ]
train_data = train_data[,c('approved','loan_amount','income','interest_rate','derived_sex')]
#train_data = train_data[complete.cases(train_data),]
test_data = test_data[,c('approved','loan_amount','income','interest_rate','derived_sex')]
#test_data = test_data[complete.cases(test_data),]

train_data$interest_rate = as.numeric(train_data$interest_rate)
test_data$interest_rate = as.numeric(test_data$interest_rate)


# Verify the dimensions
dim(train_data)
dim(test_data)
```

4. **Creating a Biased Dataset**
   
   Transform the training data to create systematic sampling bias:
   - Include female applicants with 80% probability
   - Include non-female applicants (male, joint, or undisclosed) with 20% probability
   
   This selective sampling approach simulates the challenges of obtaining truly representative data in real-world contexts.
   
```{r create_biased_data}
# Function to create biased dataset with specified sampling probabilities
create_biased_data <- function(data, female_prob = 0.8, non_female_prob = 0.2) {
  set.seed(456)
  
  # Create separate subsets for female and non-female applicants
  female_applicants <- data[((data$derived_sex == "Female")&(data$approved == 1)), ]
  non_female_applicants <- data[!(((data$derived_sex == "Female")&(data$approved == 1))), ]
  
  # Sample with specified probabilities
  female_sample <- female_applicants[sample(nrow(female_applicants), 
                                           size = round(nrow(female_applicants) * female_prob)), ]
  
  non_female_sample <- non_female_applicants[sample(nrow(non_female_applicants), 
                                                    size = round(nrow(non_female_applicants) * non_female_prob)), ]
  
  # Combine the samples to create the biased dataset
  biased_data <- rbind(female_sample, non_female_sample)
  
  # Shuffle the rows to mix the data
  biased_data <- biased_data[sample(nrow(biased_data)), ]
  
  return(biased_data)
}


biased_train_data = create_biased_data(train_data)
```

5. **Model Comparison**
   
   Fit both linear regression and LASSO models to the biased dataset.
   - Compare the resulting regression coefficients with those from the unbiased model
   - Document how sampling bias influences parameter estimates
   
```{r model_comparison}
# Create model matrix for LASSO on biased data
x_biased <- model.matrix(~ income + interest_rate + factor(derived_sex) , data = biased_train_data)
y_biased <- biased_train_data$approved

#linear regression
lm_model_biased <- glm(approved ~ income + interest_rate + factor(derived_sex) , data = biased_train_data,family = 'binomial')


#linear regression
lm_model_unbiased <- glm(approved ~ income + interest_rate + factor(derived_sex), data = train_data,family = 'binomial')




```




6. **Prediction Evaluation**
   
   Apply the biased models to the test dataset:
   - Generate loan amount predictions
   - Analyze prediction accuracy stratified by applicant sex and race
   - Identify whether certain demographic groups receive systematically better or worse estimates
   
```{r}

plot(predict(lm_model_biased,newdata = test_data,type = 'response'), predict(lm_model_unbiased,newdata = test_data,type = 'response'))
```
   
   
   
   
```{r}
summary(lm_model_biased)
summary(lm_model_unbiased)
```
   

7. **Sensitivity Analysis**
   
   Vary the 80/20 probability split used in step 4 to create datasets with different levels of bias:
   - Implement a range of sampling probabilities
   - Assess how varying degrees of sampling bias affect prediction accuracy
   - Determine the threshold at which sampling bias begins to significantly compromise model performance
   - Establish criteria for "reasonable" accuracy and identify the maximum tolerable sampling bias